---
title: "Activity: Bird Species"
author: "Your Names Here"
date: "DUE DATE "
output: html_notebook
---


# Set Up

```{r}
# Load `tidyverse` and `dcData` packages

library(tidyverse)
library(dcData)


# Load the `OrdwayBirds` data set into your RStudio environment from the `dcData` package

data("OrdwayBirds")


# Show a few commands used to inspect the data set a few different ways

?OrdwayBirds
OrdwayBirds
```

```{r}
OrdwayBirds |>
  group_by(Sex) |>
  summarize(count = n())
```

```{r}
OrdwayBirds |>
  group_by(SpeciesName) |>
  summarize(count = n())
```

# Activity

When you're finished with this activity, you will have a graph that shows what time of year various species appear at the Katherine Ordway Natural History Study Area in Inver Grove Heights, MN.

### Step 0

Before we begin, the book gives us some instructions to select a few key variables and clean up the date formatting.  Use the commands given in the book, but be sure to study the functions used so you know what is happening.  In this case, we select `SpeciesName`, `Month`, and `Day`. Then a `mutate()` command converts each variable to character, and then converts the resulting "character string" to a number.

```{r}
# Get the data table & clean up dates (see Data Computing eBook)
OrdwayBirds <-
  OrdwayBirds %>%
  select( SpeciesName, Month, Day ) %>%
  mutate( Month = as.numeric(as.character(Month)), 
          Day = as.numeric(as.character(Day)))


# Inspect resulting data
OrdwayBirds

```



### Step 1




```{r}
# unique species in original data (including mis-spellings)

SpeciesCount <- OrdwayBirds |> summarize(numOfSpecies = n_distinct(SpeciesName))
SpeciesCount # 275 w/ mis-spellings 

# unique species in the clean list
data("OrdwaySpeciesNames")
DistinctSpeciesCount <- OrdwaySpeciesNames |> summarize(numOfSpeciesCleaned = n_distinct(SpeciesNameCleaned))
DistinctSpeciesCount # 109 distinct in cleaned names set
```


### Step 2
  



```{r}
# inspect OrdwaySpeciesNames

OrdwaySpeciesNames # contains 2 variables: SpeciesName and SpeciesNameCleaned. This table appears to match various spellings of the species names to a standardized spelling of each name (Gold Finch, Goldfinch, and Golf Finch are mapped to Goldfinch, for example).


# inner join
CorrectedOrdwayBirds <- OrdwayBirds |>
  inner_join(OrdwaySpeciesNames, by = "SpeciesName", relationship = "many-to-many") |>
  select(Species = SpeciesNameCleaned, Month, Day) |>
  na.omit()

# I made some minor changes to the code snippet from the textbook. Adding a `by` clause explicitly marks that we are joining on that variable, and stating that the relationship is to `many-to-many` stops the interpreter thing from warning us every time we run this code block (we expect many-to-many since there are multiple occurrences of the various spellings in the original table)

# inspect the data after the join

CorrectedOrdwayBirds

```


### Step 3

We want to identify the top 5-6 bird species sighted at the Katherine Ordway Natural History Study Area.

```{r}
# count sightings by species in descending order

DescendingBirds <- CorrectedOrdwayBirds |>
  group_by(Species) |>
  summarize(count = n()) |>
  arrange(desc(count))


# a hint in the text book recommends displaying the top 10 species to choose our threshold



# define a sightings threshold; produce a table with only birds belonging to major species

threshold <- 1000

Majors <- DescendingBirds |>
  filter(count >= threshold)

Majors

```


### Step 4

We now want to produce a month-by-month count for each of the major species, and then construct a chart to tell the story of what time of year the various species appear.

```{r}
# prepare glyph-ready data table called `ByMonth`

ByMonth <- OrdwayBirds |>
  filter(SpeciesName %in% Majors$Species) |>
  group_by(Month, SpeciesName) |>
  summarize(count = n())

```

Now use the ggplot template below to generate the desired graph. (Remark you can also use esquisser or mplot to create a line chart instead of the barchart)



```{r}
# ggplot commands

ByMonth %>% 
  ggplot(aes(x = Month, y = count, color = SpeciesName)) + 
  geom_line()

ByMonth %>% ggplot(aes(x=Month,y=count ,fill=Species)) +
  geom_bar(stat='identity',position='stack', width=.9) 

```

According to the graph, we can answer the questions posed:

##### 1. Which species are present year-round?  

The `American Goldfinch`, `Black-capped Chickadee`, and `Slate-colored Junco` are all major species that appear to be present all year long.

##### 2. Which species are migratory, that is, primarily present in one or two seasons?

The `Field Sparrow` and `Tree Swallow` appear to be migratory, seeing a sthey appear somewhere around March and disappear sometime around November (basically not the winter).


##### 3. What is the peak month for each major species? (bullet list)

- The `American Goldfinch` peaks in October.
- The `Black-capped Chickadee` appears to peak sometime around November.
- The `Field Sparrow` peaks in May.
- The `Slate-colored Junco`, the most surveyed species in this set, primarily peaks in October with a sizable second peak occuring around April.
- The `Tree Swallow` appears to peak somewhere in late June/early July.


##### 4. Which major species are seen in good numbers for at least 6 months of the year?

While each of the species seem rather absent during the summer months, there are species that collectively are present for a total of at least 6 months over the course of a given year. The `Slate-colored Junco` is present in no small amounts up to May, making a reappearance as the summer months wind down in August/September. Proportionally, the `Black-capped Chickadee` follows a similar trend, with the main differences being that its numbers are a lot smaller than those of the `Slate-colored Junco` and, instead of disappearing in the summer months, it stays only mildly present. In fact, with the exception of the `Tree Swallow`, all of the species seem to maintain a proportionally sizable presence over the course of at least 6 months in a given year. The `Tree Swallow` thrives in June, but quickly returns to obscurity in the surrounding months. As such, it's difficult to claim that it's seen in "good numbers" for at least 6 months of the year.
